Exploring the Traveling Salesman Problem

### 2017-05-03 paper/paper.tex
update

% Packages
% ---
\usepackage[
  left=3cm,
  right=3cm,
  top=2cm,
  bottom=2cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{clrscode3e}
\usepackage{blkarray}
\usepackage{booktabs}
\usepackage{parskip}
\usepackage{mathptmx}
\usepackage{bigstrut}
\usepackage{mathptmx}
\usepackage{minted}
\usepackage[pdftex,pdfpagelabels,bookmarks,hyperindex,hyperfigures]{hyperref}
\usepackage{apacite}
\usepackage{tikz}

@@

% Document
% ---
\begin{document}
%\setlength{\abovedisplayskip}{0pt}
%\setlength{\belowdisplayskip}{0pt}
%\setlength{\abovedisplayshortskip}{0pt}
%\setlength{\belowdisplayshortskip}{0pt}

\title{Exploring the Traveling Salesman Problem}
\author{Brennan Hoeting}
\date{}
\maketitle

\section*{Introduction}
This paper will explore the Traveling Salesman Problem (TSP),
a classic NP---hard problem in the field of computer science.
We will first obtain an understanding of the problem. That
is, we will explain the problem, examine its input and output,
and show why solving this problem particularly challenging.
We will present and explain a naive brute force solution to the
problem and analyze it's efficiency.  We will present a better
method for solving the TSP, the Held---Karp algorithm, and explain
it in detail mathematically and with pseudocode.  There is a
smorgasbord of algorithms that improve over the brute force method,
but they all have their drawbacks and none of them run in a polynomial
time complexity.  We will examine a few of these methods in lesser
detail than Held---Karp and look at situations in which one algorithm
is favorable over another.

\section{Understanding the Problem}
\subsection{Overview}
The traveling salesman problem can be described as follows:
Given a set of cities, where there is a distance between
each pair of cities, find a tour (route) that begins at a
starting city, visits each \textit{destination city} exactly once,
and returns to the starting city with the minimum possible
total distance traveled.  We will refer to the tour with the
minimum possible distance traveled as the \textit{optimal tour}.

\subsection{Problem Input}
The problem input is a matrix that defines the distance
between each pair of cities.  We will denote this as
the distance matrix $D = (d_{ij})$, where $d_{ij}$ is
the distance between city $i$ and city $j$.  It is important
to note that the cost (distance) from city $i$ to city $j$ is equal
to the cost from city $j$ to city $i$. This means that our
TSP variation is \textit{undirected}, so we may use a diagnol
matrix to represent the distances.  That is, for every $d_{ij}$
where $i<j$, $d_{ij}$ is the distance between cities $i$ and $j$,
and every $d_{ij}$ where $i\geq j$, is ignored.
\par

In a \textit{directed} variation of the TSP --- such as one that
optimizes for minimum total airfair to travel a tour by plane,
where ticket prices from city $i$ to city $j$ is not necessarily
equal to the ticket price from city $j$ to city $i$ --- we would
store a cost in every value of $d_{ij}$ where $i\neq j$, rather
than storing the costs in a diagnol matrix.  This distinction,
while helpful to be aware of, minimally affects how the problem is
solved in the methods we will explore.
\par

See below for a valid input example for an undirected TSP variation.
Notice, for example, that the distance between city $1$ and city $4$ is
$d_{1,4}=29$.  If we ran into a case while implementing a TSP algorithm
where we are finding the distance between $i$ and $j$ but $j>i$, we
simply retrieve the distance from $d_{ji}$ instead of $d_{ij}$.
\par

\begin{align*}
  d_{ij}=
  \begin{blockarray}{cccccc}
    & j_1 & j_2 & j_3 & j_4 & j_5 \\
  \begin{block}{c(ccccc)}
    i_1 & \cdot & 7 & 37 & 29 & 17 \\
    i_2 & \cdot & \cdot & 31 & 32 & 34 \\
    i_3 & \cdot & \cdot & \cdot & 32 & 42 \\
    i_4 & \cdot & \cdot & \cdot & \cdot & 20 \\
    i_5 & \cdot & \cdot & \cdot & \cdot & \cdot \\
  \end{block}
  \end{blockarray}
\end{align*}

\subsection{Problem Output}
The output for the TSP is also a matrix with the dimensions
$N\times N$, where $N$ is the number of cities.  We will define the output as
the edge matrix $E = (e_{ij})$, where $e_{ij}=1$ if the optimal tour involves
traveling from city $i$ to city $j$.  $e_{ij}=0$ if meaning we will not
travel from city $i$ to city $j$ along the optimal tour.
\par

We use an edge matrix because if we had a graph $G=(V,E)$, $V_i$
would represent city $i$ and $E_{ij}$ would
represent traveling from city $i$ to city $j$ along the optimal tour.
If the graph where visualized, each verticy would be positioned at its
longitude and latitude coordinates and a segment would connect each pair
of verticies $v_i$ and $v_j$ where $e_{ij}=1$.
\par

Below is an example of a valid output for our TSP variation.  Notice that
$e_{3,4}=1$ implying our optimal tour involves traveling from city $3$ to
city $4$.  Observe that when $i<j$, just as the distance matrix $d_{ij}$,
the value in $e_{ij}$ is not used, resulting in another diagonal matrix.
Observe that $e_1$ has two columns that are $1$.  This will always be the
case for a single $e_i$ where $i$ is the city at we we begin and terminate
the tour.  In this case, and throughout the paper, the optimal tour will
always start and end on the first city.
\par

\begin{align*}
  e_{ij}=
  \begin{blockarray}{cccccc}
    & j_1 & j_2 & j_3 & j_4 & j_5 \\
  \begin{block}{c(ccccc)}
    i_1 & \cdot & 1 & 0 & 0 & 1 \\
    i_2 & \cdot & \cdot & 1 & 0 & 0 \\
    i_3 & \cdot & \cdot & \cdot & 1 & 0 \\
    i_4 & \cdot & \cdot & \cdot & \cdot & 1 \\
    i_5 & \cdot & \cdot & \cdot & \cdot & \cdot \\
  \end{block}
  \end{blockarray}
\end{align*}

\section{A Naive Brute Force Algorithm}
\subsection{Overview}
The brute force method for solving the Traveling Salesman Problem has exactly
two redeeming qualities: (1) it is quite simple, making implementation
trivial, and (2) it will produce the correct tour in every TSP instance.
Unfortunately, for instances that contain more than roughly 10 cities, the
algorithm does not finish execution in a reasonable time.  Analyzing the
pseudocode in (2.2), we can get a good idea of why this is the case.
\par
\vspace{0.5cm}

@@

\li    \Return $\proc{MAKE-EDGE-MATRIX}(\id{min-tour})$
\end{codebox}

@@

\vspace{.27cm}
\hrule
\begin{codebox}
\Procname{$\proc{TOUR-DISTANCE}(t:$ tour, $D:$ distance matrix$)$}
\li $\id{dist} \gets 0$
\li \For $\id{i} \gets 2$ to $\attrib{t}{length}$
\li   \Do
$\id{dist} \gets \id{dist} + D_{\proc{min}(t_{i-1}, t_i), \proc{max}(t_{i-1}, t_i)}$
       \End
\li    \Return $\id{dist}$
\end{codebox}
\vspace{.1cm}
\hrule
\vspace{.05cm}

If $\id{dist} < \id{min-dist}$, we update
$\id{min-dist}$ and $\id{min-tour}$.  Once the loop is finished, we
return an edge matrix to represent the optimal tour using the
$\proc{MAKE-EDGE-MATRIX}()$ procedure, which creates an edge matrix from a tour
represented as a list.
\par

@@

\section{A Better Algorithm}
\subsection{Overview}
While the brute force algorithm is simple and correct, it
is far too slow for larger inputs.  Fortunately, there are multiple
alternative TSP algorithms with a non-polynomial runtime
that is better than $O(n!)$.  We will explore in detail the
\textit{Held--Karp} algorithm, an $O(2^n)$ solution that uses
dynamic programming \cite{heldkarp}.  Then we will look at it
to the MST approximation algorithm and discuss the advantages
of each approach.

@@

This alone does not solve the original problem.  We still need to return 
the optimal tour, not just its distance.  We can find the optimal tour
by backtracing through $H$.  We define the optimal tour as $P$ and $B(v)$
as a recurrence that backtraces through $H$:

\begin{align*}
  B(v)=\begin{cases}
    B(H[v])     & \text{if } v \text{\ in\ } H \\
    \const{nil} & \text{otherwise} 
  \end{cases}
\end{align*}

We fill in the values of $P$ by calling $B((\{1,\dots,n\}, 1))$.  On each
recurrence where $v$ is in $H$, we append the second value from the tuple $v$
to $P$.  When $v$ is not in $H$ and $\const{nil}$ is returned from $B(v)$, we
append $1$ to $P$, resulting in the optimal tour.
\par

@@

\subsection{Held-Karp Pseudocode}

>>>>

% Packages
% ---
\usepackage[
  left=3cm,
  right=3cm,
  top=2cm,
  bottom=2cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{clrscode3e}
\usepackage{blkarray}
\usepackage{booktabs}
\usepackage{parskip}
\usepackage{mathptmx}
\usepackage{bigstrut}
\usepackage{mathptmx}
\usepackage{minted}
\usepackage[pdftex,pdfpagelabels,bookmarks,hyperindex,hyperfigures]{hyperref}
\usepackage{apacite}
\usepackage{tikz}

@@

% Replace outlined proof square with a black square.
\renewcommand\qedsymbol{$\blacksquare$}

% Document
% ---
\begin{document}

% Title
% ---
\title{Exploring the Traveling Salesman Problem}
\author{Brennan Hoeting}
\date{}
\maketitle

% Introduction
% ---
\section*{Introduction}
The Traveling Salesman Problem (TSP) is a classic NP---hard problem
in the field of computer science.  Like every other NP--hard problem,
it can only be solved in exponential runtime, or approximated in polynomail
runtime.  In this paper, we will obtain an understanding of the TSP problem
by explaining what it is, examining its input and output, and looking at
different algorithms that can obtain or approximate a solution.  We will
present and explain a naive brute force solution to the problem and analyze
its efficiency.  Then we will look at alternative TSP algorithms and discuess
the use cases and advantages of each approach.  We will choose one of the
algorithms and present its pseudocode, as well as detailed mathematical
notation to gain a better understanding of how it computes the solution.

\section{Understanding the Problem}
\subsection{Overview}
The traveling salesman problem can be described as follows:
Given a set of cities, where there is a distance between
each pair of cities, find a tour (route) that begins at a
\textit{starting city}, visits each \textit{destination city}
exactly once, and returns to the starting city covering the minimum
possible total distance.  We will refer to the tour with the
minimum possible distance traveled as the \textit{optimal tour}.

\subsection{Problem Input}
The problem input is a matrix that defines the distance
between each pair of cities.  We will refer to this as
a \textit{distance matrix} or \textit{cost matrix} and denote
it as $D = (d_{ij})$, where $d_{ij}$ is the distance between city
$i$ and city $j$.  If the cost (distance) from city $i$ to city $j$
is equal to the cost from city $j$ to city $i$ then the TSP variation
is \textit{undirected} and our distance matrix is symmetric.
\par

An example of a \textit{directed} variation of the TSP is one that
optimizes for minimum total airfair to travel a tour by plane,
where ticket prices from city $i$ to city $j$ is not necessarily
equal to the ticket price from city $j$ to city $i$, implying the
cost matrix is asymmetric.
\par

Most exact TSP algorithms can solve both directed and undirected variations,
while approximation TSP algorithms can typically only solve undirected
variations.
\par

See the figure below for a valid input distance matrix for an undirected
TSP variation.  Notice, for example, that the distance between city $1$
and city $4$ is $d_{1,4}=d_{4,1}=29$.  If we ran into a case while
implementing a TSP algorithm
\par

\begin{align*}
  d_{ij}=
  \begin{blockarray}{cccccc}
    & j_1 & j_2 & j_3 & j_4 & j_5 \\
  \begin{block}{c(ccccc)}
    i_1 & 0 & 7 & 37 & 29 & 17 \\
    i_2 & 7 & 0 & 31 & 32 & 34 \\
    i_3 & 37 & 31 & 0 & 32 & 42 \\
    i_4 & 29 & 32 & 32 & 0 & 20 \\
    i_5 & 17 & 34 & 32 & 20 & 0 \\
  \end{block}
  \end{blockarray}
\end{align*}

It is common to describe the input of the TSP as a complete graph.
We define the graph as $G=(V,E)$, where $v_i$ is the $i^{\text{th}}$
city and $e_ij$ is the distance between city $i$ and city $j$.  Below
is an example what such a graph could look like:

\tikzstyle{vertex}=[circle, draw, minimum size=1.2cm]
\tikzstyle{edge}=[draw=black]
\begin{tikzpicture}[shorten >=1pt, auto, node distance=7cm]
  \node[vertex](v1) at (6, 4) {$ v_1 $};
  \node[vertex](v2) at (11, 2) {$ v_2 $};
  \node[vertex](v3) at (10, -2) {$ v_3 $};
  \node[vertex](v4) at (2, -2) {$ v_4 $};
  \node[vertex](v5) at (1, 2) {$ v_5 $};

  \draw[edge] (v1) edge node{\scriptsize 7}  (v2);
  \draw[edge] (v1) edge node{\scriptsize 37} (v3);
  \draw[edge] (v1) edge node{\scriptsize 29} (v4);
  \draw[edge] (v1) edge node{\scriptsize 17} (v5);

  \draw[edge] (v2) edge node{\scriptsize 31} (v3);
  \draw[edge] (v2) edge node{\scriptsize 32} (v4);
  \draw[edge] (v2) edge node{\scriptsize 34} (v5);

  \draw[edge] (v3) edge node{\scriptsize 32} (v4);
  \draw[edge] (v3) edge node{\scriptsize 42} (v5);

  \draw[edge] (v4) edge node{\scriptsize 42} (v5);
\end{tikzpicture}

\subsection{Problem Output}
A common output for a TSP is a matrix with the dimensions $N\times N$, where
$N$ is the number of cities.  We can define the output as the edge matrix
$E = (e_{ij})$, where $e_{ij}=1$ if the optimal tour involves
traveling from city $i$ to city $j$.  $e_{ij}=0$ if meaning we will not
travel from city $i$ to city $j$ along the optimal tour.
\par

Below is an example of a valid edge matrix output for our TSP variation.
Notice that $e_{3,4}=1$ implying our optimal tour involves traveling from
city $3$ to city $4$.  Since this output is for an undirected TSP instance,
$e_{4,3}$ is also $1$.
\par

\begin{align*}
  e_{ij}=
  \begin{blockarray}{cccccc}
    & j_1 & j_2 & j_3 & j_4 & j_5 \\
  \begin{block}{c(ccccc)}
    i_1 & 0 & 1 & 0 & 0 & 1 \\
    i_2 & 1 & 0 & 1 & 0 & 0 \\
    i_3 & 0 & 1 & 0 & 1 & 0 \\
    i_4 & 0 & 0 & 1 & 0 & 1 \\
    i_5 & 1 & 0 & 0 & 1 & 0 \\
  \end{block}
  \end{blockarray}
\end{align*}

TSP algorithms can also about a graph $G=(V,E)$, where $V_i$
represents city $i$ and $E_{ij}$ represents traveling from city $i$ to
city $j$ along the optimal tour, where the weight is the distance between
the two.  If the graph where visualized, each verticy would be positioned
at its longitude and latitude coordinates and a segment would connect each pair
of verticies $v_i$ and $v_j$ where $e_{ij}=1$.  The single path in the graph
represents the optimal tour.
\par

\tikzstyle{vertex}=[circle, draw, minimum size=1.2cm]
\tikzstyle{edge}=[draw=black]
\begin{tikzpicture}[shorten >=1pt, auto, node distance=7cm]
  \node[vertex](v1) at (6, 4) {$ v_1 $};
  \node[vertex](v2) at (11, 2) {$ v_2 $};
  \node[vertex](v3) at (10, -2) {$ v_3 $};
  \node[vertex](v4) at (2, -2) {$ v_4 $};
  \node[vertex](v5) at (1, 2) {$ v_5 $};

  \draw[edge] (v1) edge node{\scriptsize 7}  (v2);
  \draw[edge] (v2) edge node{\scriptsize 31} (v3);
  \draw[edge] (v3) edge node{\scriptsize 32} (v4);
  \draw[edge] (v4) edge node{\scriptsize 42} (v5);
  \draw[edge] (v5) edge node{\scriptsize 17} (v1);
\end{tikzpicture}

Another way to represent the optimal tour is with a list
$T$, where for every $i \lte n$, $(i-1, i)$ represents
traveling from city $i-1$ to city $i$ along the optimal
tour.  This paper will primarily use this type of output.

\section{A Naive Brute Force Algorithm}
\subsection{Overview}
The brute force method for solving the Traveling Salesman Problem has exactly
two redeeming qualities: (1) it is quite simple, making implementation
trivial, and (2) it will produce the correct tour in every TSP instance.
Unfortunately, for instances that contain more than roughly 20 cities, the
algorithm does not finish execution in a reasonable time.  Analyzing the
pseudocode in (2.2), we can get a good idea of why this is the case.
\par
\vspace{0.5cm}

@@

\li    \Return $\id{min-tour}$
\end{codebox}

@@

\vspace{.27cm}
\hrule
\begin{codebox}
\Procname{$\proc{TOUR-DISTANCE}(t:$ tour, $D:$ distance matrix$)$}
\li $\id{dist} \gets 0$
\li \For $\id{i} \gets 2$ to $\attrib{t}{length}$
\li   \Do
$\id{dist} \gets \id{dist} + D_{i-1,i}$
       \End
\li    \Return $\id{dist}$
\end{codebox}
\vspace{.1cm}
\hrule
\vspace{.05cm}

If $\id{dist} < \id{min-dist}$, we update
$\id{min-dist}$ and $\id{min-tour}$.  Once the loop is finished, we
return the optimal tour.
\par

@@

\section{A Better Algorithm}
\subsection{Overview}
While the brute force algorithm is simple and correct, it
is far too slow for larger inputs.  Fortunately, there are multiple
alternative TSP algorithms with a runtime
that is better than $O(n!)$.  We will explore in detail the
\textit{Held--Karp} algorithm, an $O(n^2 2^n)$ solution that uses
dynamic programming \cite{heldkarp}.  Then we will look at the Nearest
Neighbor algorithm and Christofides algorithm, both of which are polynomial
time approximation algorithms.  We will and discuss the advantages
and drawbacks of each approach.

@@

This alone does not solve the original problem.  We still need to return 
the optimal tour, not just its distance.  We can find the optimal tour
by using an additional lookup table, $P$, that maps $(S, x)$ to the optimal
$(S-\{x\}, k)$ at each step.  We can obtain the final optimal tour by
backtracking through $P$.  We define the optimal tour as $T$ and $B(v)$ as a
recurrence that backtracks through $P$:

\begin{align*}
  B(v)=\begin{cases}
    B(P[v])     & \text{if } v \text{\ in\ } P \\
    \const{nil} & \text{otherwise} 
  \end{cases}
\end{align*}

We fill in the values of $T$ by calling $B((\{1,\dots,n\}, 1))$.  On each
recurrence where $v$ is in $P$, we append the second value in $v$ (the $k$
value) to $T$.  When $v$ is not in $H$ and $\const{nil}$ is returned
from $B(v)$, we append $1$ to $P$, resulting in the optimal tour.
\par

@@

\subsection{Held-Karp Pseudocode}
Below is the Pseudocode for the Held-Karp algorithm.  It is an iterative
dynamic programming implementation of the recurrence we defined in (3.2).
It uses two lookup tables, $C$ and $P$, for computing the optimal distances
and backtracking to obtain the optimal path, respectively.
\par

\vspace{.27cm}
\hrule
\begin{codebox}
\Procname{$\proc{HELD-KARP}(D:$ distance matrix$)$}
\li $\id{n} \gets \attrib{D}{length}$
\li $\id{C} \gets$ a lookup table that maps $(S, x)$ to the optimal cost
\li $\id{P} \gets$ a lookup table that maps $(S, x)$ to the optimal $(S-\{x\}, k)$
\li
\li \For $\id{k} \gets 2$ to $n$
\li   \Do
	$\id{C} \gets D_{1k}$
\li     $\id{P} \gets (\{\}, 1)$
      \End
\li
\li \For $\id{size} \gets 2$ to $n$
\li   \Do
	\For $\id{x} \gets 1$ to $n$
\li	  \Do
	    \If $size \isequal n$ and $x \neq 1$
\li	      \Then
		\textbf{continue}
	      \End
\li	    \For $\id{S} \gets \proc{SUBSETS}(\{2,\dots,n\}, size)$
\li	      \Do
		$\id{opt-k} \gets \const{nil}$
\li		$\id{opt-dist} \gets \infty$
\li		$\id{opt-subset} \gets \const{nil}$
\li		\For $k$ in $S-\{x\}$
\li		  \Do
		    $\id{dist} \gets \id{C}[(S-\{x\}, \id{k})] + D_{xk}$
\li		    \If $dist < \id{opt-dist}$
\li		      \Then
			$\id{opt-k} \gets k$
\li			$\id{opt-dist} \gets dist$
\li			$\id{opt-subset} \gets S$
		      \End
		  \End
\li
\li		  $\id{C}[(subset, x)] \gets opt-dist$
\li		  $\id{P}[(subset, x)] \gets (opt-subset, opt-k)$
	      \End
	  \End
      \End

\li
\li $T = []$
\li $\id{v} \gets (\{1,\dots,n\}, 1)$
\li \While $\id{v} \neq \const{nil}$
\li   \Do
	$\proc{\attrib{T}{append}}(v[2])$
\li	$\id{S} \gets \id{T}[v]$
      \End	
\li
\li \Return \id{T}
\end{codebox}
\vspace{.27cm}
\hrule

On lines 5-7, we add our base cases to the lookup tables.  In the main loop on
lines 9-25, we build the lookup tables in a bottom-up dynamic programming fashion.
Notice on lines 11 and 12 we continue when the subset size is $n$ and $x \neq 1$.
This acheives the effect of beginning our recurrence from (3.2) with $(\{1,\dots,n\}, 1)$,
ensuring we're only computing optimal paths that start at $1$.  Finally, on lines 27-31
we backtrack through our lookup table $P$ that stores the optimal $(S-\{x\}, k)$
for a given $(S, x)$, appending each optimal $k$ value to the optimal tour $T$.

\subsection{Held-Karp Runtime}
We will show that the Held-Karp algorithm is an improvement over the brute force
approach by proving the runtime of Held-Karp is $O(n^{2}2^n)$, which is better than
the runtime of brute force, $O(n!)$.  Observe from the pseudocode that our algorithm
involves computing every possible subset of $\{1,\dots,n\}$, where $n$ is the number
of cities.  Knowing this, we can prove the runtime by proving a set with $n$ elements
has $2^n$ subsets using induction.

\begin{proof}
  (Induction) For our basis step, consider that an empty set $\varnothing$ has $1$
  subset (itself).  $2^n = 2^0 = 1$, so a set with $0$ elements has $2^0$ subsets.\
  For induction, suppose a $k$-element set has $2^k$ subsets.  We will show every
  $(k+1)$-element subset has $2^(k+1)$ subsets.  Let $A=\{a_1,a_2,\dots,a_k, b\}$,
  notice it contains $k+1$ elements.  We parition the subsets of $A$ into two sets
  $B$ and $C$, where $B$ contains the subsets that contain $b$ and $C$ contains
  the subsets that don't contain $b$.  $C$ contains every subset of $A-{b}$, so it
  has $2^k$ elements.  By construction, it follows that $B$ must have the same number
  of elements as $C$, so it also has $2^k$ elements.  Since $B$ and $C$ are independent
  sets, the size of $B\union C$ is $2^k + 2^k = 2^{k+1}$.
\end{proof}
